---
title: 'Binary Logistic Regression'
author: "Mariam Walaa"
output:
  html_document: default
  pdf_document: default
---

```{r echo = FALSE, warning = FALSE, message = FALSE}
packages = c("MASS", "broom", "corrplot", "tidyverse")
install.fcn <- function(x){
  if (!require(x, character.only = TRUE)){
    install.packages(x, dependencies = TRUE)
  }
}
package.check <- lapply(packages, install.fcn)
```

```{r warning = FALSE, message = FALSE}
library(MASS)
library(broom)
library(corrplot)
library(tidyverse)
```

### Data

This data set comes from the UCI Machine Learning Repository and can be found
[here](https://archive.ics.uci.edu/ml/datasets/wine+quality). According to the data
abstract, this data set consists of red vinho verde wine samples from the north of
Portugal. The purpose of this data set is to model wine quality based on physicochemical
tests. This data set contains one integer variable and the remaining variables are
doubles. There are 1,599 red wine samples and 12 variables describing each sample. There
are no missing values in this data set. The dependent variable is wine quality.

```{r}
wine <- read.csv(file = '../data/Wine/winequality-red.csv', sep = ';')
```

```{r, fig.width = 15, fig.height = 15}
plot(wine)
```

```{r, fig.width = 20, fig.height = 20}
wine %>% 
  cor() %>%
  corrplot(type = "upper", order = "hclust")
```

```{r, fig.width = 12, fig.height = 8}
hist(wine$quality, main = "Wine Quality By Score (3-8)")
```

Most wines have a score of 5 or 6 for quality.

```{r}
wine$quality <- ifelse(wine$quality > 6, 1, 0)
```

Any wines with a score > 6 is considered to be of good quality.

```{r}
stepAIC(glm(quality ~.,
            data = wine,
            family = binomial(link = 'logit')), 
        trace = FALSE)$anova
```

The step AIC model selection algorithm suggests that the best model for predicting the
quality of wine is the model that uses fixed acidity, volatile acidity, residual sugar,
chlorides, total sulfur dioxide, density, sulphates, and alcohol to predict the quality of
wine.

```{r}
model <- glm(quality ~
               fixed.acidity +
               volatile.acidity +
               residual.sugar +
               chlorides + 
               total.sulfur.dioxide +
               density +
               sulphates +
               alcohol,
             data = wine,
             family = binomial(link = 'logit'))
```

```{r}
summary(model)
```

All independent variables in this model are statistically significant with p-values of <
0.001 indicating enough evidence to reject the null hypothesis that there is no
relationship between the independent variable and the dependent variable. By the
distribution of the deviance residuals, the residuals of the model do not appear to be
normally distributed but are close.

```{r, fig.width = 12, fig.height = 8}
plot(model, which = 4, id.n = 4)
```

There are 4 values that have a much larger Cook's distance than the rest.

```{r}
augment(model) %>%
  mutate(index = 1:n()) %>%
  top_n(4, .cooksd) %>%
  dplyr::select(index, .resid, .std.resid, .cooksd)
```

Only 1 of these values has a standardized residual greater than 3. All have very small
Cook's distances.

```{r}
round(exp(coef(model)), 3)
```

```{r}
round(exp(confint(model)), 3)
```

Most of the confidence intervals for each independent variable are not too wide with exception to sulphates.