---
title: 'Binary Logistic Regression'
author: "Mariam Walaa"
output:
  html_document: default
  pdf_document: default
---

```{r warning = FALSE, message = FALSE}
library(tidyverse)
library(MASS)
library(datasets)
library(corrplot)
library(nnet)
library(broom)
```

### Data

This data set comes from the UCI Machine Learning Repository and can be found
[here](https://archive.ics.uci.edu/ml/datasets/wine+quality). According to the data
abstract, this data set consists of red vinho verde wine samples from the north of
Portugal. The purpose of this data set is to model wine quality based on physicochemical
tests. This data set contains one integer variable and the remaining variables are
doubles. There are 1,599 red wine samples and 12 variables describing each sample. There
are no missing values in this data set. The dependent variable is wine quality.

```{r}
wine <- read.csv(file = '../data/Wine/winequality-red.csv', sep = ';')
```

```{r}
glimpse(wine)
```

```{r}
summary(wine)
```

```{r, fig.width = 20, fig.height = 20}
wine %>% 
  cor() %>%
  corrplot.mixed(order="hclust", tl.col="black")
```

```{r}
hist(wine$quality)
```

```{r}
wine$quality <- ifelse(wine$quality > 6, 1, 0)

stepAIC(glm(quality ~.,
                    data = wine,
                    family = binomial(link = 'logit')), 
        trace = FALSE)$anova
```

```{r}
model <- glm(quality ~
               fixed.acidity +
               volatile.acidity +
               residual.sugar +
               chlorides + 
               total.sulfur.dioxide +
               density +
               sulphates +
               alcohol,
             data = wine,
             family = binomial(link = 'logit'))
```

```{r}
summary(model)
```

```{r}
round(exp(coef(model)), 3)
```

```{r}
round(exp(confint(model)), 3)
```
