---
title: 'Multinomial Logistic Regression'
author: "Mariam Walaa"
output:
  html_document: default
  pdf_document: default
---

```{r echo = FALSE, warning = FALSE, message = FALSE}
packages = c("nnet", "MASS", "palmerpenguins", "corrplot", "tidyverse")
install.fcn <- function(x){
  if (!require(x, character.only = TRUE)){
    install.packages(x, dependencies = TRUE)
  }
}
package.check <- lapply(packages, install.fcn)
```

```{r warning = FALSE, message = FALSE}
library(nnet)
library(MASS)
library(palmerpenguins)
library(corrplot)
library(tidyverse)
```

## Data

This part uses the palmer penguins dataset provided by the Long Term Ecological Research
Network and including measurements for penguins located in an island in Palmer
Archipelago, such as flipper length, body mass, bill dimensions, and sex.

```{r, fig.width = 20, fig.height = 20}
penguins %>% 
  na.omit() %>%
  select(!c("species", "island", "sex")) %>%
  cor() %>%
  corrplot.mixed(order = "hclust", tl.col = "black")
```

The flipper length, bill length, and body mass are strongly positively correlated while
bill depth is negatively correlated to them all, and year of study is not correlated to
anything.

```{r}
penguins$species <- relevel(penguins$species, ref = "Gentoo")
penguins <- penguins %>%
  na.omit()
```

```{r}
summary(stepAIC(multinom(species ~.,
                         data = penguins,
                         trace = FALSE),
                trace = FALSE))$anova
```

The step AIC model selection suggests that a species can best be predicted by bill length,
bill depth, and sex.

```{r}
model <- multinom(species ~
                    bill_length_mm +
                    bill_depth_mm +
                    sex,
                  data = penguins,
                  trace = FALSE)

summary(model)
```

The multinomial regression summary provides coefficients for 2 of the 3 species, Adelie
and Chinstrap. The effect on the dependent variable for the 3rd species can be computed by
setting the independent variables to zero. Each coefficient of an independent variable can
be used to determine the change in log-odds of being a certain type of species for a 1
unit change in that independent variable.

```{r}
(1 - pnorm(abs(summary(model)$coefficients /
                 summary(model)$standard.errors), 0, 1)) * 2
```

The p-values are all > 0.05 and indicate that we cannot reject the null hypothesis that
there is no relationship between the independent variables and the dependent variable.

```{r}
round(exp(coef(model)), 3)
```

```{r}
round(exp(confint(model)), 3)
```

The confidence intervals of all the independent variables are extremely large.

```{r}
predict(model,
        penguins[,2:8],
        type = "probs") %>%
  apply(1, which.max) -> predictions

table(penguins$species, predictions)
```

Each species is predicted with 100% accuracy.