---
title: 'Principal Component Analysis'
author: "Mariam Walaa"
output:
  html_document: default
  pdf_document: default
---

```{r echo = FALSE, warning = FALSE, message = FALSE}
packages = c("factoextra", "psych", "corrplot", "tidyverse")
install.fcn <- function(x){
  if (!require(x, character.only = TRUE)){
    install.packages(x, dependencies = TRUE)
  }
}
package.check <- lapply(packages, install.fcn)
```

```{r warning = FALSE, message = FALSE}
library(factoextra)
library(psych)
library(corrplot)
library(tidyverse)
```

### Data

In this section, I will be using principal component analysis to explore 2016 census data
aggregated by City of Toronto's 140 neighborhoods. The full data set consists of more than
2000 variables regarding demographics, immigration, housing, transportation, employment,
income, ethnicity, etc. The data has been loaded into a data frame and subset to focus
only on categories about journey to work, education, labour, population type, and income.
A subset of continuous variables from these categories have been selected and raw count
variables have been normalized by the population for each neighborhood.

```{r warning = FALSE, echo = FALSE}
neighborhoods <- read.csv('../data/TorontoNeighbourhoods_2016/neighbourhood-profiles-2016-csv.csv')
neighborhoods %>%
  subset(Category == "Population"
         | Category == "Journey to work"
         | Category == "Education"
         | Category == "Income"
         | Category == "Labour"
         | Characteristic == "Population, 2016") %>%
  select(!c(1:4, 6)) -> neighborhoods

neighborhoods <- as.data.frame(t(neighborhoods))

names(neighborhoods) <- neighborhoods[1,]
neighborhoods <- neighborhoods[-1,]
neighborhoods <- tibble::rownames_to_column(neighborhoods, "neighborhood")
neighborhoods[, -1] <- apply(neighborhoods[, -1], 2,
                             function(column) as.numeric(gsub("[\\,]","", column)))

neighborhoods <- neighborhoods[,
                               c("neighborhood", "Population, 2016",
                                 "Children (0-14 years)",
                                 "Youth (15-24 years)",
                                 "Working Age (25-54 years)",
                                 "Pre-retirement (55-64 years)",
                                 "Seniors (65+ years)",
                                 "Older Seniors (85+ years)",
                                 "  Car, truck, van - as a driver",
                                 "  Car, truck, van - as a passenger",
                                 "  Public transit",
                                 "  Walked",
                                 "  Bicycle",
                                 "  Less than 15 minutes",
                                 "  15 to 29 minutes",
                                 "  30 to 44 minutes",
                                 "  45 to 59 minutes",
                                 "  60 minutes and over",
                                 "Unemployment rate (Females)",
                                 "Unemployment rate (Males)",
                                 "  Worked at home",
                                 "  No fixed workplace address",
                                 "  Worked outside Canada",
                                 "Total income: Average amount ($)",
                                 "    Employment income: Average amount ($)",
                                 "    Investment income: Average amount ($)")]
                                 
colnames(neighborhoods) <- c("neighborhood", "pop_2016",
                             "children",
                             "youth",
                             "working_age",
                             "preretired_age",
                             "senior",
                             "older_senior",
                             "drive_commute",
                             "passenger_commute",
                             "transit_commute",
                             "walk_commute",
                             "bike_commute",
                             "15min_commute",
                             "30min_commute",
                             "45min_commute",
                             "60min_commute",
                             "over_60min_commute",
                             "unemployment_female",
                             "employment_male",
                             "work_at_home",
                             "no_fixed_workplace",
                             "work_outside_canada",
                             "avg_total_income",
                             "avg_employment_income",
                             "avg_investment_income")

neighborhoods[, c("children",
                  "youth",
                  "working_age",
                  "preretired_age",
                  "senior",
                  "older_senior",
                  "drive_commute",
                  "passenger_commute",
                  "transit_commute",
                  "walk_commute",
                  "bike_commute",
                  "15min_commute",
                  "30min_commute",
                  "45min_commute",
                  "60min_commute",
                  "over_60min_commute",
                  "unemployment_female",
                  "employment_male",
                  "work_at_home",
                  "no_fixed_workplace",
                  "work_outside_canada")] <- neighborhoods[, c(
                  "children",
                  "youth",
                  "working_age",
                  "preretired_age",
                  "senior",
                  "older_senior",
                  "drive_commute",
                  "passenger_commute",
                  "transit_commute",
                  "walk_commute",
                  "bike_commute",
                  "15min_commute",
                  "30min_commute",
                  "45min_commute",
                  "60min_commute",
                  "over_60min_commute",
                  "unemployment_female",
                  "employment_male",
                  "work_at_home",
                  "no_fixed_workplace",
                  "work_outside_canada")] / neighborhoods[, "pop_2016"] 

neighborhoods <- neighborhoods %>% select(!pop_2016)

write.csv(neighborhoods, "../data/TorontoNeighbourhoods_2016/toronto_neighborhoods.csv",
          row.names = FALSE)

neighborhoods <- column_to_rownames(neighborhoods, "neighborhood")
```

### Variable Correlations

```{r, fig.width = 15, fig.height = 15}
neighborhoods %>%
  cor() %>%
  corrplot(type = 'upper', order = 'alphabet')
```


Neighborhoods where there are more people with commutes within 60 minutes are less
likely to have more people that are seniors or children and more likely to have people
that are of working age, as well as females that are unemployed. However, they are more
likely to have people who walk to work, work at home, or work outside of Canada.
Neighborhoods where there are people with higher incomes are more likely to have people in
the older demographic and less likely to have people with a commute longer than an hour or
people who commute as passengers or take transit. They are, however, more likely to work
at home or work outside of Canada.

### PCA Results

```{r}
neighborhoods %>%
  prcomp(center = TRUE, scale. = TRUE) -> neighborhoods.pca
```

### PCA Component Selection

```{r, fig.width = 10, fig.height = 6, warning = FALSE}
get_eigenvalue(neighborhoods.pca)
fviz_eig(neighborhoods.pca)
fa.parallel(neighborhoods, fa = "pc", plot = FALSE)
```

The scree plot and eigenvalues suggest taking the first 6 principal components while the
parallel analysis suggests taking the first 5 principal components.

```{r}
neighborhoods.pca$rotation[,1:5]
```

### PC Correlations

```{r, fig.width = 10, fig.height = 10}
fviz_pca_var(neighborhoods.pca,
             col.var = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)
```

Plotting the variables with the first component against the second component shows how the
variables relate to each other. The older demographic correlates negatively with short
commutes, walking, biking, and transit. The longer commutes and a non-fixed workplace
correlate negatively with income and working at home or outside of Canada. The employment
rate for males, unemployment rates for females, youth population, and 60 minute commutes
are not well-represented by the first two principal components.

```{r, fig.width = 10, fig.height = 10, warning = FALSE}
fviz_pca_ind(neighborhoods.pca,
             col.ind = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)
```

Casa Loma, Forest Hill South, Rosedale Moore Park, and Niagara are some neighborhoods that
are distant from the large cluster of neighborhoods.
