---
title: 'Principal Component Analysis'
author: "Mariam Walaa"
output:
  html_document: default
  pdf_document: default
---

```{r warning = FALSE, message = FALSE}
library(factoextra)
library(corrplot)
library(psych)
library(tidyverse)
```

### Data

In this section, I will be using principal component analysis to explore 2016 census data
aggregated by City of Toronto's 140 neighborhoods. The full data set consists of more than
2000 variables regarding demographics, immigration, housing, transportation, employment,
income, ethnicity, etc. The data has been loaded into a data frame and subset to focus
only on categories about journey to work, education, labour, population type, and income.
A subset of continuous variables from these categories have been selected and raw count
variables have been normalized by the population for each neighborhood.

```{r warning = FALSE, echo = FALSE}
neighborhoods <- read.csv('data/neighbourhood-profiles-2016-csv.csv')
neighborhoods %>%
  subset(Category == "Population"
         | Category == "Journey to work"
         | Category == "Education"
         | Category == "Income"
         | Category == "Labour"
         | Characteristic == "Population, 2016") %>%
  select(!c(1:4, 6)) -> neighborhoods

neighborhoods <- as.data.frame(t(neighborhoods))

names(neighborhoods) <- neighborhoods[1,]
neighborhoods <- neighborhoods[-1,]
neighborhoods <- tibble::rownames_to_column(neighborhoods, "neighborhood")
neighborhoods[, -1] <- apply(neighborhoods[, -1], 2,
                             function(column) as.numeric(gsub("[\\,]","", column)))

neighborhoods <- neighborhoods[ ,
                               c("neighborhood", "Population, 2016",
                                 "Children (0-14 years)",
                                 "Youth (15-24 years)",
                                 "Working Age (25-54 years)",
                                 "Pre-retirement (55-64 years)",
                                 "Seniors (65+ years)",
                                 "Older Seniors (85+ years)",
                                 "  Car, truck, van - as a driver",
                                 "  Car, truck, van - as a passenger",
                                 "  Public transit",
                                 "  Walked",
                                 "  Bicycle",
                                 "  Less than 15 minutes",
                                 "  15 to 29 minutes",
                                 "  30 to 44 minutes",
                                 "  45 to 59 minutes",
                                 "  60 minutes and over",
                                 "Unemployment rate (Females)",
                                 "Unemployment rate (Males)",
                                 "  Worked at home",
                                 "  No fixed workplace address",
                                 "  Worked outside Canada",
                                 "Total income: Average amount ($)",
                                 "    Employment income: Average amount ($)",
                                 "    Investment income: Average amount ($)")]
                                 
colnames(neighborhoods) <- c("neighborhood", "pop_2016",
                             "children",
                             "youth",
                             "working_age",
                             "preretired_age",
                             "senior",
                             "older_senior",
                             "drive_commute",
                             "passenger_commute",
                             "transit_commute",
                             "walk_commute",
                             "bike_commute",
                             "15min_commute",
                             "30min_commute",
                             "45min_commute",
                             "60min_commute",
                             "over_60min_commute",
                             "unemployment_female",
                             "employment_male",
                             "work_at_home",
                             "no_fixed_workplace",
                             "work_outside_canada",
                             "avg_total_income",
                             "avg_employment_income",
                             "avg_investment_income")

neighborhoods[, c("children",
                  "youth",
                  "working_age",
                  "preretired_age",
                  "senior",
                  "older_senior",
                  "drive_commute",
                  "passenger_commute",
                  "transit_commute",
                  "walk_commute",
                  "bike_commute",
                  "15min_commute",
                  "30min_commute",
                  "45min_commute",
                  "60min_commute",
                  "over_60min_commute",
                  "unemployment_female",
                  "employment_male",
                  "work_at_home",
                  "no_fixed_workplace",
                  "work_outside_canada")] <- neighborhoods[, c(
                  "children",
                  "youth",
                  "working_age",
                  "preretired_age",
                  "senior",
                  "older_senior",
                  "drive_commute",
                  "passenger_commute",
                  "transit_commute",
                  "walk_commute",
                  "bike_commute",
                  "15min_commute",
                  "30min_commute",
                  "45min_commute",
                  "60min_commute",
                  "over_60min_commute",
                  "unemployment_female",
                  "employment_male",
                  "work_at_home",
                  "no_fixed_workplace",
                  "work_outside_canada")] / neighborhoods[, "pop_2016"] 

neighborhoods <- neighborhoods %>% select(!pop_2016)

write.csv(neighborhoods, "toronto_neighborhoods.csv", row.names = FALSE)

neighborhoods <- column_to_rownames(neighborhoods, "neighborhood")
```

### Variable Correlations

```{r, fig.width = 15, fig.height = 15}
neighborhoods %>%
  cor() %>%
  corrplot(type = 'upper', order = 'hclust')
```

### PCA Results

```{r}
neighborhoods %>%
  prcomp(center = TRUE, scale. = TRUE) -> neighborhoods.pca
```

### PCA Component Selection

```{r, fig.width = 10, fig.height = 6, warning = FALSE}
get_eigenvalue(neighborhoods.pca)
fviz_eig(neighborhoods.pca)
fa.parallel(neighborhoods, fa = "pc", plot = FALSE)
```

### PC Correlations

```{r, fig.width = 10, fig.height = 10}
fviz_pca_var(neighborhoods.pca,
             col.var = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)
```

```{r, fig.width = 10, fig.height = 10, warning = FALSE}
fviz_pca_ind(neighborhoods.pca,
             col.ind = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)
```
