---
title: 'Spatial Autocorrelation'
author: "Mariam Walaa"
output:
  html_document: default
  pdf_document: default
---

```{r echo = FALSE, warning = FALSE, message = FALSE}
packages = c(
  # Library to load shape file into a spatial object using readOGR.
  "rgdal",
  # Library to automatically compute Moran's I using correlog.
  "pgirmess",
  # Library to:
  # - Create contiguity neighbor lists (poly2nb)
  # - Create neighbor weight lists (nb2listw)
  # - Compute the Moran's I statistic (moran.test)
  # - Estimate spatial lag models (lagsarlm)
  # - Perform Lagrange Multiplier tests (lm.LMtests)
  "spdep",
  # Library for data classification.
  "classInt",
  # Library for color schemes.
  "RColorBrewer"
  )
install.fcn <- function(x){
  if (!require(x, character.only = TRUE)){
    install.packages(x, dependencies = TRUE)
  }
}
package.check <- lapply(packages, install.fcn)
```

```{r warning = FALSE, message = FALSE}
library(rgdal)
library(pgirmess)
library(spdep)
library(classInt)
library(RColorBrewer)
```

This part will use the U.S. Census 2010 data on the Hispanic population in Salt Lake
County, estimated at the census tract level of aggregation. The shape file is projected
into U.T.M. Zone 12 coordinates (i.e. coordinates measured in meters). The shape file of
census tracts will be loaded into a spatial object below.

### Loading and Plotting

The shape file of census tracts in Salt Lake Country has 32 fields and 211 features.  A
choropleth map of the HIS_PCT variable is created below using the quantile classification
method, a sequential color scheme, and 5 classes. The HIS_PCT variable is the proportion
of Hispanic people in each census tract. The quantile classification method bins the
variable of interest into groups of equal size and is chosen for this map as it seems to
visually display more variation than Jenks natural breaks or standard deviation. The
sequential color scheme is chosen since it makes sense to use a lighter value to represent
lower proportions of Hispanic people within a census tract and a higher value to represent
higher proportions of Hispanic people within a census tract. The number of classes chosen
is 5.

```{r}
SLCCT <- readOGR(dsn = "../data/tutorial_saltlake", layer = "SLC_CT2010")
```

```{r, fig.width=10, fig.height=10}
# Define the number of classes.
nclasses <- 5

# Define the classification method. Available methods include: "equal", "sd", "quantile", "kmeans", "jenks"
class <- classIntervals(var = SLCCT$HIS_PCT, n = nclasses, style = "quantile", dataPrecision = 2)

# Set color palette.
colcode <- findColours(clI = class, pal = brewer.pal(nclasses, "PuBu"), digits = 3)

# Plot the map.
plot(SLCCT, col = colcode, border = "black", axes = T)

# Add a title.
title(main = "Proportion of Hispanic People By Census Tract, Salt Lake County 2010")

# Add a legend.
legend("bottomleft", legend = names(attr(colcode, "table")), fill = attr(colcode, "palette"), cex = 0.8)
```

```{r, fig.width=10, fig.height=6}
hist(x = SLCCT$HIS_PCT, breaks = 60, xlab = "Proportion of Hispanic People", 
     main = paste("Proportion of Hispanic People By Census Tract in Salt Lake County,", 
                "U.S. Census 2010", sep = "\n"))
summary(SLCCT$HIS_PCT)
```

The histogram of the distribution of proportion of Hispanic people in each census tract
displays a right-tailed distribution showing that majority of census tracts have 10% or
less Hispanic people within a census tract, with some outlying census tracts with more
than 50% proportion of Hispanic people in the census tract. The summary statistics table
further supports the plot of the right-tailed, non-symmetric distribution of proportions
of Hispanic people above. Both the histogram and descriptive statistics explain the
spatial patterns that are seen in the map of the Hispanic population in Salt Lake County.
Specifically, majority of the census tracts have less than 30% proportions of Hispanic
people, and this is evident in the choropleth map as only a handful of census tracts
belong to the largest class interval. In terms of spatial patterns, there seems to be a
clustering of census tracts that are of the same class interval north west and south west
of Salt Lake County, and more dispersion at the centre of the map. All of the census
tracts that are east of Salt Lake County fall into the smallest class interval of 1% to 6%
Hispanic people.

### Working with Neighbor Lists

The relationships between the census tracts can be explored in order to measure spatial
autocorrelation using neighbor graphs and weights. This is a two-step process that
consists of first choosing the neighbor criterion that needs to be satisfied and then
choosing the method to assign weights to the identified neighbor links. For the first
step, the spatial object storing the census tract geographic data has been used in the
process of creating graphs of two different neighbor criterion types (Rook and Queen).The
Queen neighbors list is expected to contain more links since the queen condition can be
met by a single shared boundary point, while the rook condition can only be met by more
than one shared point.

```{r, fig.width=10, fig.height=10}
# Retrieve coordinates from spatial object.
coords <- coordinates(SLCCT)

# Create a rook contiguity neighbor list.
SLCnb_rook <- poly2nb(pl = SLCCT, queen = FALSE)

# Create a queen contiguity neighbor list.
SLCnb_queen <- poly2nb(pl = SLCCT)

# Find differences between the two contiguity neighbor lists.
diff_nb <- diffnb(x = SLCnb_queen, y = SLCnb_rook)

# Plot the census tract boundaries.
plot(x = SLCCT, border = "grey")

# Plot the queen neighbors.
plot(x = SLCnb_queen, coords, add = TRUE, col = "black")

# Plot the differences between the rook and queens neighbor lists.
plot(x = diff_nb, coords, add = TRUE, col = "red")

# Add a title.
title(main = paste("Polygon-Generated Queen Neighbors (Black and Red)",
                 "and Polygon-Generated Rook Neighbors (Black)", sep = "\n"))

summary(card(SLCnb_rook))
summary(card(SLCnb_queen))
```

The summary statistics of the Rook and Queen neighbors shows that none of the methods have
a normal distribution, and the Queen neighbor list contains more links as expected. This
has been achieved using the helper function card which provides the cardinality of the
neighbor set. The neighbor map shows that there are many more neighbor links through the
queen condition than the rook condition. This is especially the case in areas where there
are smaller census tracts (north east and north west of the center of the map). This can
be confirmed by the number of nonzero links given by summary(SLCnb_queen), which show
there are about 300 more links using the queen condition.

### Creating Weight Matrices from Neighbor Lists

For the second step, matrices of weights have been created for the neighbor lists. These
weight matrices are contained in a listw object which is made up of the following three
attributes:

* The style of the weight matrix, which can be 'B' (binary), 'W' (row-normalized), etc.
* The input neighbors list, which is a list of values representing the number of neighbors
for each region
* The weights for each neighbor link based on the chosen style, which is a list of lists
where each sublist contains weights for that region (if using 'W' style, each region will
have different weights based on the number of neighbors the region has)

```{r}
# Create binary weight matrices
SLCnb_rook_B <- nb2listw(SLCnb_rook, style = "B")
SLCnb_queen_B <- nb2listw(SLCnb_queen, style = "B")

# Create normalized weight matrices
SLCnb_rook_W <- nb2listw(SLCnb_rook, style = "W")
SLCnb_queen_W <- nb2listw(SLCnb_queen, style = "W")

SLCnb_queen_B$weights[1]
SLCnb_queen_W$weights[1]
SLCnb_rook_B$weights[1]
SLCnb_rook_W$weights[1]
```

The weights object is a list of lists where each sub-list contains all the weights for a
census tract. For example, the first observation has 9 neighbor weights using the queen
condition and 6 neighbor weights using the rook condition. The weights for each neighbor
of a single observation or census tract are represented as 1s using the binary weighting
scheme and as decimals using the row-normalized weighting scheme. The values in the
row-normalized matrix are based on the number of neighbors for that observation.

### Global Moran's I

```{r}
# Compute Global Moran's I for each weight matrix
print(moran.test(SLCCT$HIS_PCT, listw = SLCnb_rook_B))
print(moran.test(SLCCT$HIS_PCT, listw = SLCnb_rook_W))
print(moran.test(SLCCT$HIS_PCT, listw = SLCnb_queen_B))
print(moran.test(SLCCT$HIS_PCT, listw = SLCnb_queen_W))
```

Global Moran's I is the measure of spatial autocorrelation for the entire area of study.
All Moran's I values are statistically significant so there is spatial autocorrelation for
crime levels in Columbus, Ohio. There is a higher Moran's I value for the weight matrices
using the row-normalized weighting scheme compared to the weight matrices using the binary
weighting scheme.

```{r, fig.width=10, fig.height=6}
# Compute and plot Moran's I for different distances
plot(correlog(coords, SLCCT$HIS_PCT, method = "Moran"))
```

This is a plot of global Moran's I values for the percentage of Hispanic people in each
census tract using different distances. The plot highlights the distances for which the
Moran's I statistic is statistically significant. The plot shows that a distance class
around 2000 provides the largest Moran's I statistic of above 0.6, while larger distance
classes result in less autocorrelation.
