---
title: 'Spatial Econometrics'
author: "Mariam Walaa"
output:
  pdf_document: default
  html_document: default
---

```{r warning = FALSE, message = FALSE}
library(rgdal)
library(pgirmess)
library(spdep)
library(spatialreg)
library(classInt)
library(RColorBrewer)
```

This part will use a crime data set pertaining to planning neighborhoods in Columbus, Ohio
(1980). The dependent variable of interest is CRIME, the number of residential burglaries
and vehicle thefts per 1000 households. The two independent variables are HOVAL, the
median housing value in $1000's, and INC, the median household income in \$1000's.

### Load and Describe the Data set

```{r}
columbus <- readOGR(dsn = "../data/tutorial_columbus", layer = "columbus")
```

The shape file of census tracts in Columbus, Ohio has 20 fields and 49 features.

```{r, fig.width=10, fig.height=6}

# Convert to data frame for univariate and bivariate plots below
crime_df <- as.data.frame(cbind(columbus$CRIME,columbus$INC, columbus$HOVAL))
colnames(crime_df) <- c( 'CRIME', 'INC', 'HOVAL')

# Plot pairwise correlations
plot(crime_df)

# Plot quantiles for each variable
boxplot(crime_df, main = "Boxplots of Crime, Median Income & Median Housing Value in Columbus, Ohio (1980)")
```

```{r, fig.width=10, fig.height=4}
par(mfrow = c(1,3))

# Define the number of classes.
nclasses <- 5

# Define the classification method. Available methods include: "equal", "sd", "quantile", "kmeans", "jenks"
class <- classIntervals(var = columbus$CRIME, n = nclasses, style = "quantile", dataPrecision = 0)

# Set color palette.
colcode <- findColours(clI = class, pal = brewer.pal(nclasses, "Reds"), digits = 3)

# Plot the map.
plot(columbus, col = colcode, pch = 19, axes = T, cex = 1.5)

# Add a title.
title(main = paste("Home Burglaries/Thefts Per 1000 Households",
                 "in Columbus, Ohio (1980)", sep = "\n"))

# Add a legend.
legend("topleft", legend = names(attr(colcode, "table")), fill = attr(colcode, "palette"), cex = 0.8)

# Plot a second map for median housing value
class <- classIntervals(var = columbus$HOVAL, n = nclasses, style = "quantile", dataPrecision = 0)
colcode <- findColours(clI = class, pal = brewer.pal(nclasses, "Oranges"), digits = 3)
plot(columbus, col = colcode, pch = 19, axes = T, cex = 1.5)
title(main = paste("Median Housing Value in $1000's", "in Columbus, Ohio (1980)", sep = "\n"))
legend("topleft", legend = names(attr(colcode, "table")), fill = attr(colcode, "palette"), cex = 0.8)

# Plot a third map for median household income
class <- classIntervals(var = columbus$INC, n = nclasses, style = "quantile", dataPrecision = 0)
colcode <- findColours(clI = class, pal = brewer.pal(nclasses, "Greens"), digits = 3)
plot(columbus, col = colcode, pch = 19, axes = T, cex = 1.5)
title(main = paste("Median Household Income in $1000's", "in Columbus, Ohio (1980)", sep = "\n"))
legend("topleft", legend = names(attr(colcode, "table")), fill = attr(colcode, "palette"), cex = 0.8)
```

The plot of pairwise correlations shows that crime is negatively correlated with both
median household income and median housing value, while median household income and median
housing value are positively correlated. The boxplots show that the distribution of crime
levels is normal while median household income is slightly right-skewed with an outlier
and median housing value is also right-skewed but has more outliers.

* In the crimes map, there are some clusters of census tracts belonging to the same class
interval. This is especially the case at the centre of the map for the largest class
interval, and also the case to the south for the small class interval, and also to the
middle class interval. This indicates that the proportion of crime in a census tract may
determine the proportion of crime in nearby census tracts.
* In the median housing value map, there are few clusters of census tracts belonging to
the same class interval. There are two clusters east of the map where surrounding census
tracts belong to the same class, but there is dispersion for the remainder of the map.
This indicates that the proportion of median housing value in a census tract may not
determine well what the proportion of neighboring census tracts is.
* In the median household income map, there are some clusters of census tracts belonging
to the same class interval. This is the case for multiple areas in the map, especially
south east and south west of the map for the larger class intervals. This indicates that
the median household income interval a census tract belongs to may determine the median
household income class interval of nearby census tracts.
    
### Perform Basic Spatial Statistics

```{r}

# Create a queen contiguity neighbor list.
columbusnb_queen <- poly2nb(pl = columbus)

# Create a row-normalized weight matrix.
columbusnb_queen_W <- nb2listw(columbusnb_queen, style = "W")

# Compute Global Moran's I
moran_queen_crime <- moran.test(x = columbus$CRIME, listw = columbusnb_queen_W)
moran_queen_hoval <- moran.test(x = columbus$HOVAL, listw = columbusnb_queen_W)
moran_queen_inc <- moran.test(x = columbus$INC, listw = columbusnb_queen_W)

summary(columbusnb_queen_W)
print(moran_queen_crime)
print(moran_queen_hoval)
print(moran_queen_inc)

```

Here are the summary statistics for the Queen neighbor graph:

* The average number of neighbor links for each zone is 4.816327.
* The minimum number of connections made in the neighbor list (i.e., minimum neighbors) is 2.
* The maximum number of connections made in the neighbor list (i.e., maximum neighbors) is 10.
* The Global Moran's I statistic for CRIME is 0.500188557 and it is strongly statistically significant.
* The Global Moran's I statistic for HOVAL is 0.180093114 and it is statistically significant.
* The Global Moran's I statistic for INC is 0.415628778 and it is strongly statistically significant.
* The CRIME variable is most spatially autocorrelated of all three variables with around
0.5 Moran's I autocorrelation, followed by INC with around 0.42 Moran's I autocorrelation.
HOVAL has a lower autocorrelation of around 0.18. These autocorrelation values seem
reasonable given the map patterns seen above.

### Create an OLS Model for Crime

```{r}

# Estimate an OLS model without lag
ols_model <- lm(columbus$CRIME ~ columbus$INC + columbus$HOVAL)
summary(ols_model)

# Compute Moran's I for the OLS model's residuals
moran.test(resid(ols_model), columbusnb_queen_W)
```

Here is an interpretation of the OLS model output:

* Here is the regression formula: CRIME = -1.5973\*INC -0.2739*HOVAL + 68.6190
    * For an increase of 1 unit in income, we expect to find a decrease of 1.5973 in crime.
    * For an increase of 1 unit in housing value, we expect to find a decrease of 0.2739 in crime.
* The goodness of fit (Adjusted R^2) is 0.5329 which means that around 53% of the data is
explained well by this model.
* The Moran's I measure is computed for the OLS model's residuals in order to see if the
OLS model's residuals violate the regression assumption of being independent. Based on the
Moran's I statistic, the residuals of the model are have a statistically significant
autocorrelation of 0.222109407. This means it is possible that there is a spatial
dependence between observations, so the observations may not be independently distributed.

Based on all of these results, the model may not be valid as a statistically significant
Moran's I indicates that there may be spatial dependence which will lead to biased
coefficients and underestimated standard error. Therefore, possible options for creating a
valid model given these results are estimating a spatial lag and spatial error model. The
difference between a spatial lag model and a spatial error model is that the spatial lag
model assumes that the dependent variable has spatial dependence while the spatial error
model assumes that there may be some unknown dependence in the error.

### Create a Spatially Lagged Dependent Model

This part will create a spatially lagged dependent model. A spatial lag will be created
and passed to the OLS model. This is not an appropriate method of estimation, but it will
be used to compare the differences between the OLS with lag and maximum likelihood
estimation with lag.

```{r}

# Create a new variable that contains the spatially lagged levels of crime for each zone
wy <- lag(columbusnb_queen_W, columbus$CRIME)

# Specify a new OLS model for crime as a function of the lagged crime levels variable
ols_lag_model <- lm(columbus$CRIME ~ wy + columbus$INC + columbus$HOVAL)
summary(ols_lag_model)

# Compute Moran's I for the OLS lag model's residuals
moran.test(resid(ols_lag_model), columbusnb_queen_W)

# Estimate a spatial lag model with the same predictors as above.
spatial_lag_model <- lagsarlm(columbus$CRIME ~ wy + columbus$INC + columbus$HOVAL, listw = columbusnb_queen_W)
summary(spatial_lag_model)

# Compute Moran's I for the spatial lag models' residuals
moran.test(resid(spatial_lag_model), columbusnb_queen_W)

```

|                         | OLS No Lag  |                      | OLS With Lag |          | Maximum Likelihood with Lag |            |
|-------------------------|-------------|----------------------|--------------|----------|-----------------------------|------------|
|                         | Coefficient | p-value              | Coefficient  | p-value  | Coefficient                 | p-value    |
| Constant                | 68.619      | < 0.0000000000000002 | 38.78334     | 0.000141 | 38.783341                   | 0.00002053 |
| INC                     | -1.5973     | 0.0000183            | -0.88617     | 0.017064 | -0.886175                   | 0.009785   |
| HOVAL                   | -0.2739     | 0.0109               | -0.26408     | 0.006245 | -0.264084                   | 0.002835   |
| WCRIME                  | NA          | NA                   | 0.54876      | 0.000815 | 0.548763                    | 0.008847   |
| Adj-R2                  | 0.5329      | NA                   | 0.6288       | NA       | NA                          | NA         |
| Moran's I for residuals | 0.222109407 | 0.003999             | -0.01840567  | 0.4891   | -0.01840567                 | 0.4891     |

Here is a description of the differences between the statistics of the three models above:

* The coefficients of the median housing value (HOVAL) and the median household income
(INC) are negative in all three models. These coefficients are similar for the models with
a spatial lag (OLS and Maximum Likelihood). The coefficients of the OLS without a lag are
lower. All three models have coefficient estimates that are statistically significant.
* Both OLS models (with a lag and without a lag) have an adjusted R^2 larger than 50%. The
OLS model with a lag has a larger adjusted R^2.
* The spatial autocorrelation of the residuals is around 0.22 for the OLS with no lag and
around -0.02 for the two models with a lag (both OLS and Maximum Likelihood).
* The OLS model with no lag differs from the OLS model with a lag in that the coefficients
for the model with no lag are much smaller.
* The OLS model with a lag differs from the Maximum Likelihood model with a lag in that
the coefficient estimates for the Maximum Likelihood model are more statistically
significant overall, but the coefficient estimates themselves for both models are about
the same.

An OLS model with a spatial lag is not an appropriate model. In this case, the spatial lag
model has coefficient estimates that are more statistically significant than the
coefficient estimates for the OLS with a lag.

### Create a Spatially Lagged Error Model for Crime

```{r}

# Estimate a spatial error model
error_model <- errorsarlm(columbus$CRIME ~
                            columbus$INC +
                            columbus$HOVAL,
                          listw = columbusnb_queen_W)

summary(error_model)

# Compute Moran's I for the residuals of the error model
moran.test(resid(error_model), columbusnb_queen_W)
```

A spatially lagged error model assumes that the error term has spatial dependence.
Compared to the OLS model without spatial lag, the spatial error model also has negative
coefficients but they are slightly different from the coefficients for the OLS model with
no spatial lag. Specifically, the OLS with no lag estimates a larger decrease in income
but a smaller decrease in housing value compared the spatially lagged error model. The
coefficients for the spatially lagged error model are more statistically significant
overall compared to the OLS with no lag. Compared to the spatial error model, the spatial
lag model estimates a smaller decrease in income and housing value with slightly more
statistical significance than the spatial error model. There seems to be a small
difference in the coefficients for the spatial error model and the OLS with no lag, which
may mean that there is no spatial dependence in the error term and it may not be a valid
model.

### Choose Between Several Valid Models

```{r}
# Perform Lagrange Multiplier diagnostics
lm_tests <- lm.LMtests(ols_model, listw = columbusnb_queen_W, test = "all")

# Summarize the outputs into a table
lm_results <- t(sapply(lm_tests, function(test) c(test$statistic, test$parameter, test$p.value)))
colnames(lm_results) <- c("statistic", "df", "p-value")

printCoefmat(lm_results)
```
